---
title: "R Notebook"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    extra_dependencies: caption
  header-includes:
  - \usepackage{fontspec}
  - \setmainfont{DejaVu Sans}
  - \captionsetup[table]{labelformat=empty}
  html_document:
    df_print: paged
---

```{r include=FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(knitr)
library(kableExtra)
library(rstatix)
library(coin)
library(patchwork)
library(ggpubr)
library(viridis)
source("/Users/mattparker/Dropbox/Workspace/moops/manuscript1/helper_utils.r")
knitr::opts_chunk$set(dev = "cairo_pdf")

```


```{r}
# List all CSV files in the directory
csv_files <- list.files(path = "/Users/mattparker/Dropbox/Workspace/moops/manuscript1/data/experiment_all/sample_taxon_reports", pattern = ".*.csv$", full.names = TRUE)
seq_metrics = read.csv("/Users/mattparker/Dropbox/Workspace/moops/manuscript1/data/sequencing_summary_metrics_all.csv", stringsAsFactors = FALSE)

# Process all CSV files and combine them
all_data <- bind_rows(lapply(csv_files, process_csv_file))

# Pond sample names
pond_samples = c("s9","s10","s19","s20")

# Create sample metadata using helper function
sample_metadata = create_sample_metadata(seq_metrics, pond_samples) %>%
  select(sample_name, base_sample, library_prep, experiment, water_control, infection_class, sample_type)

# Merge the taxonomy data with metadata
merged_data <- all_data %>%
  left_join(sample_metadata, by = "sample_name")

# Extract HPV data before filtering
hpv_data = merged_data %>% 
  filter(experiment == 1 | experiment == 2) %>% 
  filter(str_detect(name, regex("Alphapapillomavirus", ignore_case = TRUE)))

pcv_data = merged_data %>%
  filter((experiment == 1 & library_prep == "MSSPE")) %>% 
  filter(str_detect(name, regex("Circovirus sp.", ignore_case = TRUE)))

# Filter to include only species level (tax_level == 1) to avoid duplication
all_data_species <- merged_data %>%
  filter(tax_level == 1) %>% 
  #filter(experiment == 2 | experiment == 3) %>% 
  filter(nt_rpm >= 1) %>% # so zeros don't skew comparison
  filter(water_control == "No") %>%   # filter out Negative Controls
  filter(infection_class != "Definite") %>%   # filter out positive control samples
  add_row(hpv_data)  # but keep hpv data from exp 1 & 2
  #add_row(pcv_data) # and pcv from exp 1

# Recode values for publication
all_data_species = all_data_species %>% 
  mutate(
    sample_type = recode(sample_type, `environmental` = "wastewater"),
    library_prep = recode(library_prep, `NGS` = "mNGS")
  ) 
```


```{r}
# Assign standardized names for viral targets
name_targets = c(
  "PBoV",   # Porcine bocavirus
  "PCV",    # Porcine circovirus 2
  "PKV",    # Porcine kobuvirus / Aichivirus C
  "EV-G",   # Porcine enterovirus G
  "PTeV",   # Porcine teschovirus A
  "PAstV",  # Porcine astrovirus 4
  "PSapoV", # Porcine sapovirus
  "PSapeV", # Porcine sapelovirus
  "HPV18"   # Human papillomavirus 7/18
)

# Filter to viral category and exclude phages
all_viruses <- all_data_species %>%
  filter(category == "viruses") %>% 
  filter(is_phage == "false") %>% 
  filter(!str_detect(name, paste(c("Siphoviridae","Podoviridae","Myoviridae"), collapse = "|")))

# Add standardized name groups using helper function
all_viruses = add_viral_name_groups(all_viruses)

# Small constant for log transformation of zero RPM values
eps <- 0.1

# Filter to only the target viral species or genus
only_viral_targets = all_viruses %>% 
  group_by(base_sample, library_prep, name_group) %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) %>% # Filter to targeted viruses
  summarize(rpm_sum = sum(nt_rpm, na.rm=TRUE), .groups="drop") %>% # sum rPM for each target
  pivot_wider(
    names_from = library_prep,
    values_from = rpm_sum
  ) %>%
  mutate(
    MSSPE = replace_na(MSSPE, 0.1), # Set NAs to 0.1 for non-detections
    mNGS   = replace_na(mNGS, 0.1),   # Set NAs to 0.1 for non-detections
    rpm_enrichment = (`MSSPE` + eps)/(`mNGS` + eps),
    log2FC = log2(rpm_enrichment),
    name_group = fct_reorder(name_group, log2FC, .fun = mean, .desc = TRUE)
  ) %>% 
  arrange(desc(log2FC))

# calculate average enrichment for all targets
mean_enrich_overall = only_viral_targets %>% 
  summarise(
    mean_log2FC = mean(log2FC),
    median_foldchange = median(rpm_enrichment),
    mean_foldchange = mean(rpm_enrichment))

mean_enrich_overall

# calculate mean so we can add to the boxplot
mean_df <- only_viral_targets %>%
  group_by(name_group) %>%
  summarise(mean_value = mean(rpm_enrichment, na.rm = TRUE), median_value = median(rpm_enrichment, na.rm = TRUE)) 

# plot fold change across all samples
alltarget_boxplot <- ggplot(only_viral_targets, aes(x = name_group, y = rpm_enrichment)) +
  geom_boxplot(width = 0.8, outlier.shape = 21, outlier.fill = "white", alpha = 0.6, width = 0.6, color = "black") +
  scale_y_log10() +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2, color = "#1976D2") +  # Professional blue
  geom_hline(yintercept = 0, linetype = "dashed") +
  #dotted mean line
  geom_segment(
    data = mean_df,
    aes(x = as.numeric(name_group) - 0.3,
        xend = as.numeric(name_group) + 0.3,
        y = mean_value, yend = mean_value),
    inherit.aes = FALSE,
    linetype = "dashed", color = "black", linewidth = 0.6
  ) +
  geom_text(data = mean_df, aes(x = name_group, y = mean_value, label = paste0("μ ", round(mean_value, 1))), vjust = -1, color = "black", size = 3.2, fontface = "italic") +
  scale_x_discrete() +  # adds space between boxes for labels
  labs(
    title = "",
    x = "",
    y = "Fold enrichment (log10)",
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    #axis.text.x = element_text(angle = 45, hjust = 1),
  )

alltarget_boxplot
```


Viral Enrichment w/o primers from experiment 1
```{r}
# Create separate dataset for experiment 1 analysis
merged_data_exp1 <- all_data %>%
  left_join(sample_metadata, by = "sample_name")

# Filter to include only species level (tax_level == 1) to avoid duplication
all_data_species_exp1 <- merged_data_exp1 %>%
  filter(tax_level == 1) %>% 
  filter(experiment == 1) %>%  # Only experiment 1
  filter(water_control == "No") %>%   # Exclude negative controls
  filter(infection_class != "Definite") %>%    # Exclude positive controls
  filter(nt_rpm >= 1) # rpm at least 1

# Filter to viral category and exclude phages
all_viruses_exp1 <- all_data_species_exp1 %>%
  filter(category == "viruses") %>% 
  filter(is_phage == "false")

# Add standardized name groups using helper function
all_viruses_exp1 = add_viral_name_groups(all_viruses_exp1) %>% 
  filter(!name_group %in% c("HPV18","PCV"))

# Filter to only the target viral species or genus
only_viral_targets_exp1 = all_viruses_exp1 %>% 
  group_by(base_sample, library_prep, name_group) %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) %>% # Filter to targeted viruses
  summarize(rpm_sum = sum(nt_rpm, na.rm=TRUE), .groups="drop") %>% 
  pivot_wider(
    names_from = library_prep,
    values_from = rpm_sum
  ) %>%
  mutate(
    MSSPE = replace_na(MSSPE, 0.1), # Set NAs to 0 for non-detections
    NGS   = replace_na(NGS, 0.1),   # Set NAs to 0 for non-detections
    rpm_enrichment = (`MSSPE` + eps)/(`NGS` + eps),
    log2FC = log2(rpm_enrichment),
    name_group = fct_reorder(name_group, log2FC, .fun = mean, .desc = TRUE)
  ) %>% 
  arrange(desc(log2FC))

# calculate average enrichment for all targets
mean_enrich_overall_exp1 = only_viral_targets_exp1 %>% 
  filter(name_group != "PToV*") %>% 
  summarise(mean_log2FC = mean(log2FC), median_foldchange = median(rpm_enrichment))

mean_enrich_overall_exp1

# calculate mean so we can add to the boxplot
mean_df_exp1 <- only_viral_targets_exp1 %>%
  group_by(name_group) %>%
  summarise(mean_value = mean(rpm_enrichment, na.rm = TRUE)) 

# plot fold change across all samples
alltarget_boxplot_exp1 <- ggplot(only_viral_targets_exp1, aes(x = name_group, y = rpm_enrichment)) +
  geom_boxplot(width = 0.8, outlier.shape = 21, outlier.fill = "white", alpha = 0.6, width = 0.6, color = "black") +
  scale_y_log10() +
  geom_jitter(width = 0.1, alpha = 0.6, size = 1.5, color = "#E67E22") +  # Professional orange
  geom_hline(yintercept = 0, linetype = "dashed") +
  # dotted mean line
  geom_segment(
    data = mean_df_exp1,
    aes(x = as.numeric(name_group) - 0.3,
        xend = as.numeric(name_group) + 0.3,
        y = mean_value, yend = mean_value),
    inherit.aes = FALSE,
    linetype = "dotted", color = "red", linewidth = 0.6
  ) +
  geom_text(data = mean_df_exp1, aes(x = name_group, y = mean_value, label = paste0("μ ", round(mean_value, 1))), vjust = -1, color = "grey40", size = 3.2, fontface = "italic") +
  scale_x_discrete() +  # adds space between boxes for labels
  labs(
    title = "Viral Enrichment (Experiment 1)",
    x = "",
    y = "Fold enrichment",
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
  )

alltarget_boxplot_exp1
```

Combined Enrichment Plot
```{r}
# Combine data from both experiments for unified visualization
combined_targets <- bind_rows(
  only_viral_targets %>% mutate(experiment_group = "Targeted Enrichment"),
  only_viral_targets_exp1 %>% mutate(experiment_group = "Off-Target Enrichment")
) %>% 
  filter(!name_group %in% c("HPV18","PCV"))

# Create combined boxplot with faceting
combined_enrichment_plot <- ggplot(combined_targets, aes(x = name_group, y = rpm_enrichment, fill = experiment_group)) +
  geom_boxplot(width = 0.7, outlier.shape = 21, outlier.fill = "white", alpha = 0.7, position = position_dodge(0.8)) +
  scale_y_log10() +
  geom_point(aes(color = experiment_group), 
             position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8),
             alpha = 0.6, size = 1.5) +
  scale_fill_manual(values = c("Targeted Enrichment" = "#3C8DBC", "Off-Target Enrichment" = "#E67E22")) +
  scale_color_manual(values = c("Targeted Enrichment" = "#3C8DBC", "Off-Target Enrichment" = "#E67E22")) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "",
    x = "",
    y = "Fold enrichment (log10)",
    fill = "Dataset",
    color = "Dataset"
  ) +
  theme_bw() +
  theme(
    #axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top",
    legend.title = element_blank()
  )

combined_enrichment_plot
```

Final rPM Enrichment Tables for Manuscript
```{r, results='asis'}
cat("\\captionsetup{labelformat=empty}")

# final table for each viral target
viral_target_enrichment_summary = only_viral_targets %>% 
  group_by(name_group) %>% 
  summarise(
    rpm_mngs = round(mean(mNGS, na.rm = TRUE),2), # mean
    rpm_msspe = round(mean(MSSPE, na.rm = TRUE),2), #mean
    median_enrichment = round(median(rpm_enrichment, na.rm = TRUE),2),
    mean_enrichment = round(mean(rpm_enrichment, na.rm = TRUE),2),
    log_fold_enrichment = round(mean(log2FC, na.rm = TRUE),2),
    iqr = paste0(
      round(quantile(rpm_enrichment, 0.25, na.rm = TRUE), 1),
      "–",
      round(quantile(rpm_enrichment, 0.75, na.rm = TRUE), 1)
    ),
    n_pairs = n(),
    p_value = {
      # ensure both vectors are aligned and remove any NA pairs
      paired_data <- na.omit(data.frame(mngs = mNGS, msspe = MSSPE))
      
      if(nrow(paired_data) > 1) {
        round(wilcox.test(paired_data$mngs,
                    paired_data$msspe,
                    alternative = "two.sided",
                    paired = TRUE)$p.value, 3)
      } else {
        NA_real_
      }
    },
  effect_r = {
      paired_data <- na.omit(data.frame(mngs = mNGS, msspe = MSSPE))
      if(nrow(paired_data) > 1) {
        long_df <- paired_data %>%
          pivot_longer(cols = c(mngs, msspe), names_to = "prep", values_to = "value") %>%
          mutate(prep = factor(prep, levels = c("mngs","msspe")))
        round(wilcox_effsize(long_df, value ~ prep, paired = TRUE)$effsize, 3)
      } else NA_real_
    }
  ) %>% 
  arrange(desc(median_enrichment))


# Make table pretty
viral_target_table <- viral_target_enrichment_summary %>%
  rename(
    "Virus" = name_group,
    "rPM (mNGS)" = rpm_mngs,
    "rPM (MSSPE)" = rpm_msspe,
    "Median Fold Change" = median_enrichment,
    "Mean Fold Change" = mean_enrichment,
    "IQR" = iqr,
    "p-value" = p_value,
    "effect size (r)" = effect_r
  ) %>% 
  select(-n_pairs, -log_fold_enrichment) %>% 
  kable(
    format = "latex",
    booktabs = TRUE,
    caption = "Enrichment across all paired samples per viral target",
    digits = 3,
    align = "lrrr",   
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    latex_options = c("hold_position","scale_down"), 
    position = "center",
    full_width = FALSE,
    font_size = 9
  ) %>%
  row_spec(0, bold = TRUE) %>%  
  collapse_rows(columns = 1, latex_hline = "major") 

viral_target_table
```

Proportion of samples with viral targets
```{r}

count_target_viruses = all_viruses %>% 
  filter(!str_detect(sample_name, "PC")) %>% 
  group_by(sample_name, name_group) %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) %>% 
  summarize(count = sum(n(), na.rm=TRUE), .groups="drop")

virus_sample_per_targets = count_target_viruses %>% 
  group_by(name_group) %>% 
  summarize(
    perc_of_samples = n_distinct(sample_name) / n_distinct(.$sample_name)
    ) %>% 
  arrange(desc(perc_of_samples))
```


GENOME COVERAGE


```{r, results='asis'}
cat("\\captionsetup{labelformat=empty}")

# Folder containing coverage CSVs
cov_dir <- "/Users/mattparker/Dropbox/Workspace/moops/manuscript1/data/experiment_all/coverage_tables"

# Read all CSV files
cov_files <- list.files(cov_dir, pattern = "\\.csv$", full.names = TRUE)

# Function to compute breadth and mean depth from one coverage file
calc_cov_stats <- function(file) {
  df <- read.csv(file)
  tibble(
    file = basename(file),
    virus = str_replace(file, "_s.*|_pc.*", ""),
    base_sample = str_remove(str_extract(basename(file), "(?<=_)(s\\d+|pc\\d+)"), "_"),
    prep = ifelse(str_detect(file, "_m"), "MSSPE", "NGS"),
    sample_type = if_else(base_sample %in% pond_samples, "environmental", "slurry"),
    ref_length = nrow(df),
    breadth_1x = mean(df$Coverage > 0) * 100,  # Percentage of genome covered at ≥1x depth
    mean_depth = mean(df$Coverage)  # Mean coverage depth across genome
  )
}

# Calculate coverage for all files
cov_summary <- map_dfr(cov_files, calc_cov_stats)

# Pivot wider for paired comparison
paired_tbl <- cov_summary %>%
  select(virus, base_sample, prep, breadth_1x, mean_depth) %>%
  pivot_wider(
    names_from = prep,
    values_from = c(breadth_1x, mean_depth),
    names_glue = "{.value}_{prep}"
  ) %>%
  mutate(
    enrichment_breadth = breadth_1x_MSSPE - breadth_1x_NGS,  # Absolute percentage point increase
    enrichment_depth   = mean_depth_MSSPE - mean_depth_NGS,  # Absolute depth increase
    virus = recode(virus,
      `astrov` = "PAstV",
      `etvg` = "EV-G",
      `hpv18` = "HPV18",
      `pbov` = "PBoV",
      `pcv` = "PCV",
      `sapel` = "PSapeV",
      `tev` = "PTeV",
      `pkv` = "PKV",
      `sapo` = "PSapoV",
    ),
    virus = fct_reorder(virus, enrichment_breadth, .fun = mean, .desc = TRUE)
  ) %>% 
  arrange(desc(enrichment_breadth))

# Summary table (1 row per metric)
wilcox_summary = paired_tbl %>% 
  replace_na(list(breadth_1x_MSSPE = 0, breadth_1x_NGS = 0, enrichment_breadth = 0)) %>% # Some samples had reads but no coverage
  group_by(virus) %>%
  summarise(
    mean_ngs = round(mean(breadth_1x_NGS, na.rm = TRUE),2),
    mean_msspe = round(mean(breadth_1x_MSSPE, na.rm = TRUE),2),
    percent_enrichment = round(mean_msspe - mean_ngs,2),
    iqr = paste0(
      round(quantile(breadth_1x_MSSPE - breadth_1x_NGS, 0.25, na.rm = TRUE), 1),
      "–",
      round(quantile(breadth_1x_MSSPE - breadth_1x_NGS, 0.75, na.rm = TRUE), 1)
    ),
    n_pairs = n(),
    p_value = if (n() > 1) {
      test <- wilcox.test(
        breadth_1x_MSSPE,
        breadth_1x_NGS,
        paired = TRUE,
        alternative = "two.sided",
        exact = FALSE
      )
      round(test$p.value, 3)
    } else {
      NA_real_
    },
    effect_r = {
      paired_data <- na.omit(data.frame(ngs = breadth_1x_NGS, msspe = breadth_1x_MSSPE))
      if(nrow(paired_data) > 1) {
        long_df <- paired_data %>%
          pivot_longer(cols = c(ngs, msspe), names_to = "prep", values_to = "value") %>%
          mutate(prep = factor(prep, levels = c("ngs","msspe")))
        round(wilcox_effsize(long_df, value ~ prep, paired = TRUE)$effsize, 3)
      } else NA_real_
    },
    .groups = "drop"
  )


# Make table pretty
wilcox_summary_kbl = wilcox_summary %>%
  mutate(
    virus = recode(virus,
      `astrov` = "PAstV",
      `etvg` = "EV-G",
      `hpv18` = "HPV18",
      `pbov` = "PBoV",
      `pcv` = "PCV",
      `sapel` = "PSapeV",
      `tev` = "PTeV",
      `pkv` = "PKV",
      `sapo` = "PSapoV",
    )
  ) %>% 
  rename(
    "Virus" = virus,
    "% coverage (mNGS)" = mean_ngs,
    "% coverage (MSSPE)" = mean_msspe,
    "Increase (%)" = percent_enrichment,
    "IQR" = iqr,
    "p-value" = p_value,
    "effect size (r)" = effect_r
  ) %>% 
  select(-n_pairs) %>% 
  kable(
    format = "latex",  # or "latex" if using LaTeX
    booktabs = TRUE,
    caption = "",
    digits = 3,
    align = "lrrr",   # adjust as needed: text left, numbers right
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    latex_options = c("hold_position","scale_down"), # polished journal look
    position = "center",
    full_width = FALSE,
    font_size = 9
  ) %>%
  row_spec(0, bold = TRUE) %>%  # strong header
  collapse_rows(columns = 1, latex_hline = "major") # if repeating target names
```

```{r}
# calculate mean so we can add to the boxplot
mean_cov_df <- paired_tbl %>%
  group_by(virus) %>%
  summarise(mean_value = mean(enrichment_breadth, na.rm = TRUE)) 
  
# Create the boxplot for each targeted virus
all_target_cov_plot = ggplot(paired_tbl, aes(x = virus, y = enrichment_breadth)) +
  geom_boxplot(width = 0.6, outlier.shape = 21, outlier.fill = "white", color = "black") +
  geom_jitter(width = 0.1, alpha = 0.6, size = 2, color = "#7B1FA2") +  # Professional green
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_segment(
    data = mean_cov_df,
    aes(x = as.numeric(virus) - 0.3,
        xend = as.numeric(virus) + 0.3,
        y = mean_value, yend = mean_value),
    inherit.aes = FALSE,
    linetype = "dashed", color = "black", linewidth = 0.6
  ) +
  geom_text(data = mean_cov_df, aes(x = virus, y = mean_value, label = paste0("μ ", round(mean_value, 1))), vjust = -1, color = "black", size = 3.2, fontface = "italic") +
  labs(
    x = "",
    y = "Increase in genome coverage (%)",
    title = ""
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

all_target_cov_plot
```

```{r}
# Overall Enrichment (non-paired)

# All viruses & enrichment 
overall_tbl = all_viruses %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) %>% # Filter to targeted viruses
  group_by(name_group, library_prep) %>% 
  summarize(rpm_mean = sum(nt_rpm, na.rm = TRUE), .groups = "drop")

overall_tbl_summary = all_viruses %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) %>% # Filter to targeted viruses
  group_by(base_sample, name_group, library_prep) %>% 
  summarize(rpm_mean = sum(nt_rpm, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = library_prep,
    values_from = rpm_mean
  ) %>%
  mutate(
    MSSPE = replace_na(MSSPE, 0.1), # Set NAs to 0 for non-detections
    mNGS   = replace_na(mNGS, 0.1),   # Set NAs to 0 for non-detections
    rpm_enrichment = (`MSSPE` + eps)/(`mNGS` + eps),
    log2FC = log2(rpm_enrichment),
    name_group = fct_reorder(name_group, log2FC, .fun = mean, .desc = TRUE)
  ) %>% 
  arrange(desc(log2FC))

# Calculate average enrichment overall
median(overall_tbl_summary$log2FC)
round(quantile(overall_tbl_summary$log2FC, 0.25, na.rm = TRUE), 2)
round(quantile(overall_tbl_summary$log2FC, 0.75, na.rm = TRUE), 2)

# median enrichment overall
median(overall_tbl_summary$rpm_enrichment)
# mean
mean(overall_tbl_summary$rpm_enrichment)
# IQR
round(quantile(overall_tbl_summary$rpm_enrichment, 0.25, na.rm = TRUE), 2)
round(quantile(overall_tbl_summary$rpm_enrichment, 0.75, na.rm = TRUE), 2)

# Paired Wilcoxon test across all samples
wilcox_res <- wilcox.test(
  overall_tbl_summary$mNGS,
  overall_tbl_summary$MSSPE,
  paired = TRUE,
  alternative = "two.sided"
)

pval <- wilcox_res$p.value

log2fc_plot = ggplot(overall_tbl_summary, aes(x = "All paired target detections (n=103)", y = rpm_enrichment)) +
  geom_boxplot(width = 0.4, fill = "gray90", color = "black", outlier.shape = 21, outlier.fill = "white") +
  scale_y_log10() +
  geom_jitter(width = 0.1, alpha = 0.6, size = 1.5, color = "#1976D2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 1, y = max(overall_tbl_summary$rpm_enrichment, na.rm = TRUE) * 1.1,
           label = paste0("Wilcoxon p = ", signif(pval, 3)),
           fontface = "italic", size = 4) +
  labs(
    x = "",
    y = "Fold enrichment (log10)",
    title = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position = "none",
  )

log2fc_plot

# get overall genome coverage enrichment - use wilcox_summary table which includes NA as a non detection
overall_cov_tbl = wilcox_summary %>% 
  group_by(virus) %>% 
  summarize(cov_change_mean = mean(percent_enrichment), .groups = "drop") # change to depth if you want

mean(overall_cov_tbl$cov_change_mean)
round(quantile(overall_cov_tbl$cov_change_mean, 0.25, na.rm = TRUE), 2)
round(quantile(overall_cov_tbl$cov_change_mean, 0.75, na.rm = TRUE), 2)

# Is the different in coverage values significant?
wilcox.test(
  paired_tbl$breadth_1x_NGS,
  paired_tbl$breadth_1x_MSSPE,
  paired = TRUE,
  alternative = "two.sided" 
)

# Paired Wilcoxon test across all samples
wilcox_res_cov <- wilcox.test(
  paired_tbl$breadth_1x_NGS,
  paired_tbl$breadth_1x_MSSPE,
  paired = TRUE,
  alternative = "two.sided" 
)

pval_cov <- wilcox_res_cov$p.value

# summarize for boxplot
overall_cov_tbl_2 = cov_summary %>% 
  group_by(virus, prep) %>% 
  summarize(cov_change_mean = mean(breadth_1x, na.rm = TRUE), .groups = "drop")

# Boxplot of mean_nt_rpm by library_prep
cov_plot = ggplot(paired_tbl, aes(x = "All paired target detections (n=103)", y = enrichment_breadth)) +
  geom_boxplot(width = 0.4, fill = "gray90", color = "black", outlier.shape = 21, outlier.fill = "white") +
  geom_jitter(width = 0.1, alpha = 0.6, size = 1.5, color = "#7B1FA2") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 1, y = 70,
           label = paste0("Wilcoxon p = ", signif(pval_cov, 3)),
           fontface = "italic", size = 4) +
  scale_y_continuous(
    limits = c(-25, 75),
    breaks = seq(-25, 75, 25)
  ) +
  labs(
    x = "",
    y = "Genome coverage increase (%)",
    title = ""
  ) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position = "none",
  )

cov_plot

```

```{r}
horizontal_space <- plot_spacer() + plot_layout(widths = c(0.05))

combined_boxplots = (log2fc_plot | horizontal_space | cov_plot) +
  plot_layout(ncol = 3, widths = c(1.5, 0.5, 1.5))


combined_boxplots
```

Comparing Likelihood of detecting targets by Lib Prep or Sample Type
```{r}
library(purrr)

# filter to targets
filter2targets = all_viruses %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) 

# get tables to create matrix
all_targets = unique(filter2targets$name_group)
all_samples = unique(filter2targets$sample_name)
all_preps = unique(filter2targets$library_prep)
all_types = unique(filter2targets$sample_type)

# create library prep matrix + detected column
expanded <- expand_grid(
  sample_name = all_samples,
  library_prep = all_preps,
  name_group = all_targets
) %>%
  left_join(filter2targets %>% mutate(detected = 1), 
            by = c("sample_name", "library_prep", "name_group")) %>%
  mutate(detected = ifelse(is.na(detected), 0, detected))

# do mcnemars test
results_libprep <- expanded %>%
  group_by(name_group, sample_name) %>%
  summarise(
    mNGS = detected[library_prep == "mNGS"],
    MSSPE = detected[library_prep == "MSSPE"],
    .groups = "drop"
  ) %>%
  group_by(name_group) %>%
  summarise(
    mcnemar = list(mcnemar.test(table(mNGS, MSSPE))),
    p_value = mcnemar[[1]]$p.value,
    # Compute discordant counts
    b = sum(mNGS == 0 & MSSPE == 1),  # detected by MSSPE only
    c = sum(mNGS == 1 & MSSPE == 0),  # detected by NGS only

    # Odds ratio (b/c)
    odds_ratio = ifelse(c == 0, NA, b / c),
    log_or = log(odds_ratio),
    se_log_or = sqrt(1 / b + 1 / c),
    ci_lower = exp(log_or - 1.96 * se_log_or),
    ci_upper = exp(log_or + 1.96 * se_log_or),
    .groups = "drop"
  )

results_libprep
```
Overall Odds Ratio
```{r}
# Sum discordant counts across all viruses
overall_b <- sum(results_libprep$b, na.rm = TRUE)  # MSSPE only
overall_c <- sum(results_libprep$c, na.rm = TRUE)  # NGS only

# Construct 2x2 table for McNemar
overall_tbl <- matrix(
  c(NA, overall_b, overall_c, NA),  # we only need discordant counts for OR/p
  nrow = 2,
  byrow = TRUE,
  dimnames = list(NGS = c("0","1"), MSSPE = c("0","1"))
)

# Run McNemar test
# Note: mcnemar.test expects all 4 cells; can set concordant counts to 0
overall_tbl[1,1] <- 0  # neither detected
overall_tbl[2,2] <- 0  # both detected

overall_test <- mcnemar.test(overall_tbl)

# Overall odds ratio
overall_or <- ifelse(overall_c > 0, overall_b / overall_c, NA)
log_or <- log(overall_or)
se_log_or <- sqrt(1/overall_b + 1/overall_c)
ci_lower <- exp(log_or - 1.96 * se_log_or)
ci_upper <- exp(log_or + 1.96 * se_log_or)

overall_results <- tibble(
  odds_ratio = overall_or,
  log_or = log_or,
  se_log_or = se_log_or,
  ci_lower = ci_lower,
  ci_upper = ci_upper,
  p_value = overall_test$p.value
)

overall_results
```
Genus Odds Ratios
```{r}

```

Sanity checking raw rPM values for targeted viruses
```{r}
sanity_tbl = all_viruses %>% 
  group_by(base_sample, library_prep, name_group, experiment) %>% 
  filter(str_detect(name_group, regex(paste0("\\b(", paste(name_targets, collapse="|"), ")\\b"), ignore_case = TRUE))) %>% 
  summarize(rpm_sum = sum(nt_rpm, na.rm=TRUE), .groups="drop") %>% 
  # Create column names combining experiment and library prep
  mutate(exp_lib = paste0("exp", experiment, "_", tolower(library_prep))) %>%
  # Group by name_group and pivot wider
  group_by(name_group, exp_lib) %>%
  summarize(nt_rpm = sum(rpm_sum, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = exp_lib,
    values_from = nt_rpm,
    values_fill = 0  # Fill missing values with 0
  ) %>%
  # Reorder columns to have name_group first, then exp1_msspe, exp1_ngs, exp2_msspe, etc.
  select(name_group, sort(tidyselect::peek_vars()[-1]))

```

EXPORT ALL FIGURES AND TABLES FTW
```{r export_paths, include=FALSE}
fig_dir <- "/Users/mattparker/Dropbox/Workspace/moops/manuscript1/msspe_enrichment/figures"
table_dir <- "/Users/mattparker/Dropbox/Workspace/moops/manuscript1/msspe_enrichment/tables"
```

```{r export_artifacts, include=FALSE}
# Use shared helpers from helper_utils.R to export plots and tables
ensure_export_dirs(fig_dir = fig_dir, table_dir = table_dir)

# Plots
if (exists("alltarget_boxplot")) {
  export_plot_png(alltarget_boxplot, fig_dir, "alltarget_boxplot.png", width = 9, height = 5)
}
if (exists("combined_enrichment_plot")) {
  export_plot_png(combined_enrichment_plot, fig_dir, "combined_enrichment_plot.png", width = 10, height = 5)
}
if (exists("all_target_cov_plot")) {
  export_plot_png(all_target_cov_plot, fig_dir, "all_target_cov_plot.png", width = 9, height = 5)
}
if (exists("log2fc_plot")) {
  export_plot_png(log2fc_plot, fig_dir, "overall_enrichment_boxplot.png", width = 5, height = 5)
}
if (exists("cov_plot")) {
  export_plot_png(cov_plot, fig_dir, "overall_coverage_boxplot.png", width = 5, height = 5)
}
if (exists("combined_boxplots")) {
  export_plot_png(combined_boxplots, fig_dir, "combined_boxplots.png", width = 12, height = 5)
}

# Tables (CSV)
if (exists("viral_target_enrichment_summary")) {
  export_table_csv(viral_target_enrichment_summary, table_dir, "viral_target_enrichment_summary.csv")
}
if (exists("wilcox_summary")) {
  export_table_csv(wilcox_summary, table_dir, "genome_coverage_enrichment_summary.csv")
}
if (exists("results_libprep")) {
  export_table_csv(results_libprep, table_dir, "detection_odds_results_by_target.csv")
}
if (exists("overall_results")) {
  export_table_csv(overall_results, table_dir, "overall_detection_odds_results.csv")
}
if (exists("sanity_tbl")) {
  export_table_csv(sanity_tbl, table_dir, "target_virus_rpm_sanity_table.csv")
}

# Export a LaTeX-styled table to PNG 
if (exists("viral_target_table")) {
  export_table_png(
    latex_table = viral_target_table,
    table_dir = table_dir,
    output_basename = "viral_target_enrichment_table"
  )
}
if (exists("wilcox_summary_kbl")) {
  export_table_png(
    latex_table = wilcox_summary_kbl,
    table_dir = table_dir,
    output_basename = "wilcox_summary_kbl"
  )
}
```

